Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-28 submission/manuscript.Rmd
shigeki discussion edits

Using the list of mouse germline-enriched genes curated above, we found `r nrow(epigerm)` were misexpressed in male _Kdm5c_-KO EpiLCs. We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

@@

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were only enriched in KDM5C-bound promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for genes bound by KDM5C only in PNCs. We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. KDM5C binding around the TSS of many germline genes is evident in EpiLCs, but absent in PNCs (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E).

 In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%). However, KDM5C was bound to the promoter at 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX). Representative examples of KDM5C-bound and unbound germline DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including enrichment of meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.

 One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs ().

@@

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
We first characterized the misexpression of tissue-enriched genes within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not expressed within testis somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis. We demonstrated that while _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. 
Germline-enriched mRNAs can become aberrantly expressed in _Kdm5c_-KO cells indirect of KDM5C, as although KDM5C is enriched at germline gene promoters in EpiLCs, it is only bound to a subset of germline-enriched DEGs. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages. Furthermore, these data provide molecular footholds that can then be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. KDM5C may therefore negatively regulate totipotency, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023]. However, misexpression of _Dazl_ in _Kdm5c_-KO ESCs was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023] and, unlike _Dazl_, KDM5C does not bind to the _Dux_ promoter.

Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs also expressed _Dazl_, they did not express 2-cell specific genes. Misexpression of germline genes in _Kdm5c_-KO EpiLCs could indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Altogether, these data suggest _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs and the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.

Although eggs and sperm employ the same transcriptional programs for shared functions, e.g. PGC formation,  meiosis, and genome defense, some germline genes are sex specific. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex. However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. X chromosome inactivation (XCI) defects could explain why _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004] and loss of _Kdm5c_ impairs XCI[@samantaActivationXistEvolutionarily2022]. However, female germline DEGs were not significantly biased towards the X chromosome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.

@@

Unlike the previously characterized germline gene suppressors that place repressive histone modifications[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes the active histone mark histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. Yet, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Although dispensible in ESCs, KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at CGIs for select germline gene promoters. KDM5C's shifting role in germline gene repression during the transition from na&#239;ve to primed pluripotency indicates the requirement of catalytic and non-catlytic gene regulatory mechanisms can change depending upon the locus and developmental stage.

- DNAme and CpG islands
	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
	- We found only 30% of germline-enriched genes have CGIs, but the majority of CGI-free germline genes still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes.
	- CGIs highly determine KDM5C recruitment to germline gene promoters.
			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms

The above work provides the mechanistic foundation for KDM5C's repression of germline-specific transcription and offers novel insight into how dysregulation of _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain. Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022]. _APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008]. KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_. Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. Ectopic germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and can drive brain turmor formation in _Drosophila_[@janicEctopicExpressionGermline2010], indicating their dysregulation may promote genome instability and cellular de-differentiation. Intriguingly, some models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation However, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.

>>>>

Using the list of mouse germline-enriched genes curated above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs. We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

@@

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were only enriched in KDM5C-bound promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for genes bound by KDM5C only in PNCs. We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. KDM5C binding around the TSS of many germline genes is evident in EpiLCs, but absent in PNCs (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E). In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%). However, KDM5C was bound to the promoter at 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX). Representative examples of KDM5C-bound and unbound germline DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including enrichment of meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.

@@

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
We first characterized the misexpression of tissue-enriched genes within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not expressed within testicular somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis. We demonstrated that while _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, as KDM5C is only bound to a subset of germline-enriched DEGs. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages. Furthermore, these data provide molecular footholds that can then be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. Based on the de-repression of _Dazl_ and _Zscan4c_ in _Kdm5c_-KO ESCs, KDM5C was thought to promote the 2-cell-to-ESC transition[[@guptaGenomewideScreenReveals2023]. Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs also expressed _Dazl_, they did not express 2-cell specific genes. Misexpression of many germline genes in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Altogether, these data suggest _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs and the 2-cell-like state observed in _Kdm5c_-KO ESCs likely reflect KDM5C's primary role in germline gene repression.

Although eggs and sperm employ the same transcriptional programs for shared functions, e.g. PGC formation,  meiosis, and genome defense, some germline genes are sex specific. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex. However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022]. Sex differences in germline gene suppression may be a consequence of females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.

Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].

- H3K4me3 and CpGme typically do not colocalize
	- https://pubmed.ncbi.nlm.nih.gov/17334365/
	- https://www.nature.com/articles/s41594-017-0013-5
	- https://pubmed.ncbi.nlm.nih.gov/23664763/
	- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2718496/
	- https://pubmed.ncbi.nlm.nih.gov/19581485/

- CpGme was imparied at some germline genes, including ones expressed throughout life
-  Some regions, such as Dazl, were able to compensate for KDM5C loss, even though it gained H3K4me2 in EpiLCs. 
	- Multiple repressive mechanisms
We found _Kdm5c_-KO EpiLCs and the amygdala accumulate H3K4me2/3 around the TSS of germline genes, potential indicating KDM5C's cataltyic activity is necessary at later points in development.

KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as

as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at CGIs for select germline gene promoters. Our study therefore indicates the requirement of catalytic and non-catlytic gene regulatory mechanisms can change depending upon the locus and developmental stage.

- DNAme and CpG islands
	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
	- We found only 30% of germline-enriched genes have CGIs, but the majority of CGI-free germline genes still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes.

- Combine germline gene list, CGI, RFX information together. 
	- Benefit of the list is finding CGI difference and kdm5c indirect mechanisms of de-repression
    - List of genes enables us to propose this model
    - CGI positive meiosis/germline formation regulators turned on → these turn on CGI-negative late stage regulators → These promote downstream dysregulation long term

- CGIs highly determine KDM5C recruitment to germline gene promoters.
			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms

@@

The above work provides the mechanistic foundation for KDM5C-mediated repression of tissue and germline-specific genes. However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain. Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022]. _APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008]. KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_. Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. Ectopic testicular germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017], including brain turmors in _Drosophila_ and mammals[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012], indicating their dysregulation may promote genome instability and cellular de-differentiation. Intriguingly, some models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders with similar erosion of the germline versus soma boundary. Further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
